---
title: "Melanoma BCR repertoire analysis"
author: "Anna Obraztsova, Mikhail Shugay"
date: '6 June 2017'
output: pdf_document
---

```{r, fig.height=8, message=FALSE}
library(dplyr)
library(ggplot2)
library(reshape2)
library(plyr)

load('clones.rda')
codes <- read.table('SKCM-codes-samples-for-patients-tss.txt', header=F, sep="\t")
colnames(codes) <- c('patient_id', 'c1', 'c2', 'c3', 'tissue')
codes$tissue <- revalue(codes$tissue, c("Distant Metastasis" = "D.M.",
                                          "Primary Tumor" = "P.T.",
                                          "Regional Cutaneous or Subcutaneous Tissue (includes satellite and in-transit metastasis)" = "R.T.", "Regional Lymph Node" = "L.N."))
codes$tissue = as.factor(ifelse(as.character(codes$tissue) == "", NA, as.character(codes$tissue)))

meta.1 = read.table("full_metadata.txt", header = T, sep ="\t")

meta = meta.1 %>%
  dplyr::select(Code, Dead, OS.corrected, OS.uncorrected, IGH.clonality, IGH.clonality.all, IGH.coverage.by.MiXCR, Ratio.IGG1.to.IGH.by.MiXCR)

colnames(meta) = c("patient_id", "Dead", "OS.corrected", "OS.uncorrected", "IGH.clonality", "IGH.clonality.all", "IGH.coverage.by.MiXCR", "Ratio.IGG1.to.IGH.by.MiXCR")

meta = merge(codes, meta)

clones = merge(clones, meta)
```

```{r}
clones$low_cov = is.na(clones$IGH.clonality)
cc = clones %>% 
  group_by(patient_id, tissue, Dead, OS.corrected, OS.uncorrected,
           IGH.clonality, IGH.clonality.all, IGH.coverage.by.MiXCR, Ratio.IGG1.to.IGH.by.MiXCR, low_cov) %>%
  dplyr::summarise(trees = n(), max.tree.size = max(nodes),
                   nodes = sum(nodes), 
                   max.tree.diameter = max(diameter), 
                   max.tree.leaves = max(leaves),
                   max.tree.branching = max(ifelse(is.finite(branching), branching, 0)),
                   max.mutations = max(total.mut)) %>%
  ungroup
```

When filtering low-coverage samples, the difference is in overall number of trees and clonotypes, and tree diameter (in favour of LN of course)

```{r}
cc1 = cc %>% 
  filter(!low_cov) %>% # filter all samples with low coverage - cannot estimate clonality, etc there
  dplyr::select(tissue, patient_id, trees, nodes, 
                max.tree.diameter, max.tree.size, 
                max.tree.leaves, max.tree.branching, max.mutations) %>%
  melt(id.vars = c('tissue', 'patient_id'))

ggplot(cc1, aes(x = tissue, y = value)) + 
  geom_boxplot() + coord_flip() +
  facet_wrap(~variable, scales="free_x") +
  theme_bw() + #scale_y_log10() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

for (v in unique(cc1$variable)) {
  print(v)
  cc.sub = subset(cc1, variable == v)
  a1 = aov(value~tissue,cc.sub)
  print(summary(a1))
  print(TukeyHSD(a1))
}
```

### Survival analysis

```{r}
library(party)
library(survival)

ctree_opts = ctree_control(testtype = "Univariate", minbucket = 45)#mincriterion = 0.9)

make_tree = function(formula, df, ...) {
  dfct = ctree(formula, data = df %>% filter(!is.na(OS.corrected)), controls = ctree_opts)
  print(dfct)
  plot(dfct, ...)
}
```

Survival analysis using conventional MiXCR-derived metrics

```{r}
make_tree(Surv(OS.corrected, Dead) ~ tissue +
            IGH.clonality + IGH.coverage.by.MiXCR + Ratio.IGG1.to.IGH.by.MiXCR, 
            cc, "All tissues")
```

Survival analysis for different tissues including tree-based parameters

```{r}
make_tree(Surv(OS.corrected, Dead) ~ tissue +
            IGH.clonality + IGH.coverage.by.MiXCR + Ratio.IGG1.to.IGH.by.MiXCR +
            max.tree.diameter + max.tree.size +
            max.tree.leaves + max.tree.branching + max.mutations, 
            cc, "All tissues")

make_tree(Surv(OS.corrected, Dead) ~ 
            IGH.clonality + IGH.coverage.by.MiXCR + Ratio.IGG1.to.IGH.by.MiXCR +
            max.tree.diameter + max.tree.size +
            max.tree.leaves + max.tree.branching + max.mutations, 
            cc %>% filter(tissue != "L.N."), "Non-lymph node")

make_tree(Surv(OS.corrected, Dead) ~ 
            IGH.clonality + IGH.coverage.by.MiXCR + Ratio.IGG1.to.IGH.by.MiXCR +
            max.tree.diameter + max.tree.size +
            max.tree.leaves + max.tree.branching + max.mutations, 
            cc %>% filter(tissue == "L.N."), "Lymph node")
```

### R:S ratio

> TODO: merge with trees, stop codon to stop codon mutations.. - where are they from, perhaps should filter all non-functional

```{r, fig.height=4, fig.width=3}
library(tidyr)
library(reshape2)

load('shm.rda')

s1 <- s %>% group_by(patient_id, LN, mut.type) %>%
  dplyr::summarise(n = n()) %>%
  dcast(patient_id + LN ~ mut.type, value.var = "n") %>%
  mutate(RS_ratio = R/S)

s1 = merge(s1, meta)
```

```{r}
ggplot(s1, aes(x = LN, y = RS_ratio)) + geom_boxplot()

print(t.test(s1$RS_ratio ~ s1$LN)$p.value)

s2 <- s %>% group_by(patient_id, LN, tree.id, child) %>%
  dplyr::summarise(mut.num = n()) %>%
  group_by(patient_id, LN, tree.id) %>%
  dplyr::summarise(mut.num = sum(mut.num)/(n()+1))

ggplot(s2, aes(x = LN, y = mut.num)) + geom_boxplot()

print(t.test(s2$mut.num ~ s2$LN)$p.value)
```

