---
title: "shm_analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analysis of substitution type and frequency

Load preprocessed data

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(data.table)
library(scales)

load("sp_new_new.Rda")
df = data.table(df) %>%
  mutate(region = as.character(region)) %>%
  filter(type == "RNA") %>%
  mutate(region = factor(region, c("FR1","CDR1","FR2","CDR2","FR3","CDR3","FR4"))) %>%
  mutate(mutation.type = ifelse(as.character(from.aa) == as.character(to.aa), "S", "R")) %>%
  group_by(proj, type, sample, replica) %>%
  mutate(weight2 = 1/n())
```

Check if we observe well-documented increase in replacement:synonimic hypermutation ratio in CDR regions:

```{r}
p5=ggplot(df %>% mutate(weight2 = weight2/sum(weight2))) + 
  geom_histogram(binwidth = 1, aes(x=pos.aa, weight=weight2, fill=region)) +
  #geom_density(adjust=2, aes(x=pos.aa, weight=weight, fill=region), position = "stack", color = NA) +
  geom_density(binwidth = 1, aes(x=pos.aa, weight=weight2, linetype = mutation.type)) +
  scale_y_continuous("Hypermutation density", expand = c(0,0)) +
  scale_x_continuous("Position in IG, AA", expand = c(0,0)) +
  scale_linetype("SHM type") +
  scale_fill_brewer("Region", palette = "Paired") +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        legend.position = c(0.75, 0.81),
        legend.direction = "horizontal")
p5
```

Relative hypermutation burden by region for old and young

```{r}
df.shm.share = df %>%
  group_by(proj, sample, region) %>%
  summarise(count = n()) %>%
  group_by(sample) %>%
  mutate(share = count/sum(count))
```

```{r}
df.shm.1 = df %>%
  group_by(proj, sample, region, clonotype.id) %>%
  summarise(shms = n())

ggplot(df.shm.1, aes(x = shms, group = proj, color = proj)) +
  stat_ecdf() +
  scale_y_continuous("") +
  xlab("") +
  scale_fill_brewer("Age", palette = "Set1") +
  facet_wrap(~region, scales = "free", ncol=4) +
  theme_bw() +  
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        legend.position = c(0.90, 0.25),
        strip.background = element_blank())

dt.p = data.table()
for (r in unique(df.shm.share$region)) {
  tt = t.test(share~proj, df.shm.share %>% filter(region == r))
  p = tt$p.value
  dt.p = rbind(dt.p,
               data.table(region = r, p.adj = p))
}
dt.p$p.adj = p.adjust(dt.p$p.adj, method="BH")
print(dt.p)
```

```{r}
ggplot(df.shm.share, aes(x = proj, group = proj, y = share, fill = proj)) +
  geom_boxplot(width = 0.5) +
  scale_y_continuous("Share of SHMs", label = percent) +
  xlab("") +
  scale_fill_brewer("Age", palette = "Set1") +
  facet_wrap(~region, scales = "free", ncol=4) +
  theme_bw() +  
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        legend.position = c(0.90, 0.25),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        strip.background = element_blank())

dt.p = data.table()
for (r in unique(df.shm.share$region)) {
  tt = t.test(share~proj, df.shm.share %>% filter(region == r))
  p = tt$p.value
  dt.p = rbind(dt.p,
               data.table(region = r, p.adj = p))
}
dt.p$p.adj = p.adjust(dt.p$p.adj, method="BH")
print(dt.p)
```

Replacement to silent ratio

```{r}
df.shm.rs = df %>%
  group_by(proj, sample, region, mutation.type) %>%
  summarise(count = n()) %>%
  group_by(proj, sample, region) %>%
  summarise(rs = sum(c(0,count[which(mutation.type == "R")])) / sum(c(0,count[which(mutation.type == "S")])))
```

```{r}
ggplot(df.shm.rs, aes(x = proj, group = proj, y = rs, fill = proj)) +
  geom_boxplot(width = 0.5) +
  scale_y_continuous("R:S ratio") +
  xlab("") +
  scale_fill_brewer("Age", palette = "Set1") +
  facet_wrap(~region, scales = "free", ncol=4) +
  theme_bw() +  
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        legend.position = c(0.90, 0.25),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        strip.background = element_blank())

dt.p = data.table()
for (r in unique(df.shm.rs$region)) {
  tt = t.test(rs~proj, df.shm.rs %>% filter(region == r))
  p = tt$p.value
  dt.p = rbind(dt.p,
               data.table(region = r, p=p,p.adj = p))
}
dt.p$p.adj = p.adjust(dt.p$p.adj, method="BH")
print(dt.p)
```

Overall statistics

```{r}
df.shm.share.s = df %>%
  group_by(proj, sample, stat="# SHMs") %>%
  summarise(value = n())

df.shm.rs.s = df %>%
  group_by(proj, sample, mutation.type) %>%
  summarise(count = n()) %>%
  group_by(proj, sample, stat="R:S ratio") %>%
  summarise(value = sum(c(0,count[which(mutation.type == "R")])) / sum(c(0,count[which(mutation.type == "S")])))

tmp = rbind(df.shm.share.s,
             df.shm.rs.s)

p6=ggplot(tmp,
       aes(x = proj, y = value, fill = proj))+
geom_boxplot(width = 0.5) +
  scale_y_continuous("") +
  xlab("") +
  scale_fill_brewer("Age", palette = "Set1") +
  facet_wrap(~stat, scales = "free") +
  theme_bw() +  
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        strip.background = element_blank())
p6

dt.p = data.table()
for (ss in unique(tmp$stat)) {
  tt = t.test(value ~ proj, tmp %>% filter(stat == ss))
  p = tt$p.value
  dt.p = rbind(dt.p,
               data.table(stat = ss, p=p))
}
print(dt.p)
```
Substitution patterns

```{r}
df.sign = df %>%
  mutate(mutation.signature = paste(from.nt, to.nt, sep = ">"))
```

R:S ratio across different substitution patterns at nucleotide level

```{r}
df.sign.rs = df.sign %>%
  group_by(mutation.signature, mutation.type) %>%
  summarise(count = n()) %>%
  group_by(mutation.signature) %>%
  summarise(rs = count[which(mutation.type == "R")] / count[which(mutation.type == "S")])

p7=ggplot(df.sign.rs, aes(x = mutation.signature, y = rs, fill = mutation.signature)) +
  geom_bar(stat="identity", color="black") + 
  geom_hline(yintercept = 1, linetype="dashed") + 
  scale_fill_brewer("Signature", palette = "Paired") + 
  scale_y_continuous("R:S ratio", breaks=0:8) +
  xlab("") + 
  theme_bw() +  
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank())
p7
```

Share of different substitution patterns compared between young and old. Age-related difference observed for certain substitution patterns aross hypermutations.

```{r}
sign.dict = data.frame(mutation.signature = c("A>C","A>G","A>T","C>A","C>G","C>T","G>A",
                                              "G>C","G>T","T>A","T>C","T>G"),
                       mutation.signature.rep = c("A>C,T>G","A>G,T>C","A>T,T>A","C>A,G>T",
                                                  "C>G,G>C","C>T,G>A","C>T,G>A","C>G,G>C",
                                                  "C>A,G>T","A>T,T>A","A>G,T>C","A>C,T>G"))

df.sign.s = df.sign %>%
  group_by(proj, sample, mutation.signature) %>%
  summarise(count = n()) %>%
  group_by(sample) %>%
  mutate(freq = count / sum(count))
```

```{r}
df.sign.s$mutation.signature = factor(df.sign.s$mutation.signature,
                                      levels = with(df.sign.s %>% group_by(mutation.signature) %>% 
                                                      summarise(freq = sum(freq)), 
                                                    mutation.signature[order(-freq)]))

p8=ggplot(df.sign.s, aes(x=mutation.signature, y = freq, fill = proj)) +
  geom_boxplot() + 
  ylab("") + xlab("") +
  scale_fill_brewer("Age", palette = "Set1") + 
  ylab("Share of hypermutations") +
  theme_bw() +  
  theme(panel.grid.major.y = element_blank(), 
        panel.grid.minor = element_blank(),
        legend.position = c(0.8, 0.8))
p8

a = aov(freq ~ proj * mutation.signature, df.sign.s)
summary(a)

dt.p = data.table()

for (ms in unique(df.sign.s$mutation.signature)) {
  tt = t.test(freq ~ proj, df.sign.s %>% filter(mutation.signature == ms))
  p = tt$p.value
  dt.p = rbind(dt.p,
               data.table(mutation.signature = ms, p=p))
}

dt.p$p.adj = p.adjust(dt.p$p, method = "BH")
print(dt.p %>% arrange(p.adj))
```

```{r}
aa.classes = data.table(aa = strsplit("I,V,L,F,C,M,A,W,G,T,S,Y,P,H,N,D,Q,E,K,R", ",")[[1]],
                         hydrop = c(rep("hydrophobic", 8), rep("neutral", 6), 
                                    rep("hydrophilic", 6)))

df.aachange = df %>%
  merge(aa.classes %>% mutate(from.aa = aa, from.value = hydrop) %>% select(from.aa, from.value)) %>%
  merge(aa.classes %>% mutate(to.aa = aa, to.value = hydrop) %>% select(to.aa, to.value))
```

```{r}
hydrop_toint = function(x) {
  ifelse(x == "hydrophobic", 1, ifelse(x == "neutral", 0, -1))
}

dt.aachange.s = df.aachange %>%
  group_by(sample, proj, from.value, to.value) %>%
  summarise(count = n()) %>%
  group_by(sample, proj) %>%
  mutate(freq = count / sum(count),
         hydrop.change = hydrop_toint(to.value) - hydrop_toint(from.value)) %>%
  group_by(hydrop.change) %>%
  mutate(freq2 = freq / length(unique(paste(from.value, to.value))))

p9=ggplot(dt.aachange.s, aes(x=hydrop.change, group=paste(hydrop.change, proj), fill = proj, y= freq2)) +
  geom_boxplot() +
  scale_fill_brewer("Age", palette = "Set1") + 
  ylab("Share of hypermutations") +
  xlab("Hydropathy change (hydrophobic>hydrophilic) .. (hydrophilic>hydrophobic)") +
  theme_bw() +
  theme(aspect=1,
        legend.position = c(0.85,0.85),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank())
p9

dt.p = data.table()

for (hc in unique(dt.aachange.s$hydrop.change)) {
  tt = t.test(freq ~ proj, dt.aachange.s %>% filter(hydrop.change == hc))
  p = tt$p.value
  dt.p = rbind(dt.p,
               data.table(hydrop.change = hc, p=p))
}

dt.p$p.adj = p.adjust(dt.p$p, method = "BH")
print(dt.p %>% arrange(p.adj))
```

```{r}
ggsave("figures/p5.pdf", p5, width = 10, height = 5)
ggsave("figures/p6.pdf", p6, width = 5, height = 5)
ggsave("figures/p7.pdf", p7, width = 5, height = 5)
ggsave("figures/p8.pdf", p8, width = 8, height = 5)
ggsave("figures/p9.pdf", p9, width = 7, height = 5)
```
