---
title: "Distribution, RS ratio and patterns of novel SHM"
author: "Mikhail Shugay, Anna Obraztsova"
date: "17/04/2017"
output:
  pdf_document: default
  html_document: default
---

## Analysis of substitution type and frequency

Load preprocessed data

```{r, warning=FALSE, message=FALSE}
library(plyr)
library(ggplot2)
load("sp_new_new.Rda")
df <- subset(df, type == "RNA")
df$region <- factor(df$region, c("FR1","CDR1","FR2","CDR2","FR3","CDR3"))
df$mutation.type <- ifelse(as.character(df$from.aa) == as.character(df$to.aa), "S", "R")
df <- ddply(df, .(proj, type, sample, replica), transform, 
            weight = total.clonotypes / sum(total.clonotypes))
```

Check if we observe well-documented increase in replacement:synonimic hypermutation ratio in CDR regions:

```{r}
ggplot(df) + 
  geom_bar(aes(x=pos.aa, weight=weight, fill=region)) +
  geom_density(aes(x=pos.aa, weight=weight, linetype = mutation.type)) +
  ylab("Hypermutation density") +
  xlab("Position in IG, AA") +
  scale_fill_brewer(palette = "Set1") +
  theme_bw()
```

## Comparative analysis

Summarize by region and type, compute frequencies and R:S ratio

```{r}
df.1 <- ddply(df, .(proj, sample, type, replica, mutation.type, region), summarize,
              count = sum(total.clonotypes))

df.1 <- ddply(df.1, .(proj, sample, type, replica), transform,
              freq = count / sum(count))

df.1 <- ddply(df.1, .(proj, sample, type, replica, region), transform,
              ratio = count / (sum(count) - count))
```

Fraction of errors in each region

```{r}
ggplot(df.1, aes(x=proj, y = freq)) + geom_boxplot() + 
  facet_wrap(~region, scales = "free_y") + 
  xlab("") + ylab("Share of hypermutations") +
  theme_bw()
```

R:S ratio varies greatly by region

```{r}
ggplot(subset(df.1, mutation.type == "R"), aes(x=proj, y = ratio)) + 
  geom_boxplot() + 
  facet_wrap(~region, scales = "free_y") + 
  xlab("") + ylab("R:S ratio") +
  theme_bw()
```

Overall R:S ratio is higher in young

```{r}
df.2 <- ddply(df, .(proj, sample, type, replica, mutation.type), summarize,
              count = sum(total.clonotypes))

df.2 <- ddply(df.2, .(proj, sample, type, replica), transform,
              ratio = count / (sum(count) - count))

ggplot(subset(df.2, mutation.type == "R"), aes(x=proj, y = ratio)) + 
  geom_boxplot() + 
  xlab("") + ylab("R:S ratio") +
  theme_bw()

t.test(ratio ~ proj, subset(df.2, mutation.type == "R"))
```

Role of age factor can be deduced using generalized linear model for replacement hypermutation probability (binomial family). Probability of replacement hypermutations is increased by $7 \pm 1\%$ in young compared to old ($P < 10^{-6}$)

```{r}
df$R <- ifelse(df$mutation.type == "R", 1, 0)
fit <- glm(R ~ region + proj - 1, df, family = binomial())

summary(fit)
```

## Substitution patterns

R:S ratio across different substitution patterns at nucleotide level

```{r}
df$mutation.signature <- paste(df$from.nt, df$to.nt, sep = ">")

df.3 <- ddply(df, .(mutation.signature, mutation.type), summarize, count = sum(total.clonotypes))
df.3 <- ddply(df.3, .(mutation.signature), transform, ratio = count / (sum(count) - count))
df.3 <- subset(df.3, mutation.type == "R")

ggplot(df.3, aes(x = mutation.signature, y = ratio, fill = mutation.signature)) +
  geom_bar(stat="identity", color="black") + 
  geom_hline(yintercept = 1, linetype="dashed") + 
  scale_fill_brewer(palette = "Paired") + 
  scale_y_continuous("R:S ratio", breaks=0:8) +
  xlab("") + 
  theme_bw()
```

Share of different substitution patterns compared between young and old. Age-related difference observed for certain substitution patterns aross hypermutations.

```{r}
df.4 <- ddply(df, .(proj, sample, type, replica, mutation.signature), summarize,
              count = sum(total.clonotypes))

df.4 <- ddply(df.4, .(proj, sample, type, replica), transform,
              share = count / sum(count))

ggplot(df.4, aes(x = mutation.signature, group = interaction(mutation.signature, proj), 
                 y = share, fill = proj)) +
  geom_boxplot() + scale_y_log10(limits = c(0.01, 0.5)) +
  scale_fill_brewer("", palette = "Set1") + 
  xlab("") + ylab("Share of hypermutations") +
  theme_bw()

a <- aov(share ~ mutation.signature * proj, df.4)
summary(a)
```

Hydropathy change patterns observed at amino acid level. More hydrophilic -> hydrophobic amino acid hypermutations in young compared to old.

```{r}
aa.classes <- data.frame(aa = strsplit("I,V,L,F,C,M,A,W,G,T,S,Y,P,H,N,D,Q,E,K,R", ",")[[1]],
                         hydrop = c(rep("hydrophobic", 8), rep("neutral", 6), 
                                    rep("hydrophilic", 6)))

aa.classes$hydrop <- factor(aa.classes$hydrop, c("hydrophobic", "neutral", "hydrophilic"))

df.5 <- ddply(df, .(proj, sample, type, replica, from.aa, to.aa), summarize,
              count = sum(total.clonotypes))

df.5 <- merge(df.5, aa.classes, by.x = "from.aa", by.y = "aa")
df.5 <- merge(df.5, aa.classes, by.x = "to.aa", by.y = "aa")
#df.5$signature <- paste(df.5$hydrop.x, df.5$hydrop.y, sep = ">")

df.5 <- ddply(df.5, .(proj, sample, type, replica, hydrop.x, hydrop.y), summarize,
              count = sum(count))

df.5 <- ddply(df.5, .(proj, sample, type, replica), transform,
              share = count / sum(count))

ggplot(df.5, aes(x=proj, group=proj, y=share, fill = proj)) + geom_boxplot() + 
  facet_grid(hydrop.x~hydrop.y) + scale_y_log10() +
  scale_fill_brewer(palette = "Set1") + 
  xlab("") + ylab("Share of hypermutations") +
  theme_bw()

hydrop_toint <- function(x) {
  ifelse(x == "hydrophobic", 1, ifelse(x == "neutral", 0, -1))
}

df.5$hydrop.change <- with(df.5, hydrop_toint(hydrop.y) - hydrop_toint(hydrop.x))

ggplot(df.5, aes(x=hydrop.change, group = interaction(hydrop.change, proj), y=share, 
                 fill=proj)) + 
  geom_boxplot() + scale_y_log10() +
  xlab("Hydropathy change (hydrophobic>hydrophilic) .. (hydrophilic>hydrophobic)") +
  ylab("Share of hypermutations") +
  scale_fill_brewer("", palette = "Set1") + 
  theme_bw()


a <- aov(share ~ hydrop.x : hydrop.y + hydrop.x : hydrop.y : proj, df.5)
summary(a)

#not enough 'x' observations
#t.test(share ~ proj, 
#       subset(df.5, hydrop.y == "hydrophobic" & hydrop.x == "hydrophilic"))
```