---
title: "Somatic hypermutations signatures"
author: "Anna Obraztsova, Mikhail Shugay"
date: "4/24/2017"
output: pdf_document
---

## Exploratory data analysis

Load all our data, plot relative substitution frequencies. Here is some summary of what we currently have:

* We have flu (``flu``) and yellow fewer (``YF``) vaccination time-courses which track plasma (``P``) and memory (``M``) B-cells. 
* We also have ``old`` and ``young`` donors vaccinated against yellow fewer, ``P`` cells only and no controls unfortunately.
* Raji cell line (``raji``) and data from Steele 2009 (``steele``) are included for reference.

Note that we work with relative fractions of subsitutions which is computed as follows. Let the number of substitutions from base $B_i$ to base $B_j$ be $\#\left(B_i \rightarrow B_j\right)$, the absolute substitution frequncy is then $F_{ij} = \#\left(B_i \rightarrow B_j\right) / \#B_i$ where $\#B_i$ is the total number of occurences of base $B_i$ in a sample of sequences. The relative frequncy is given by normalizing all $F_{ij}$ to $\sum_{ij} F_{ij} = 1$ (to $100\%$), i.e. $f_{ij} = F_{ij} / \sum_{lk} F_{lk}$.

Also note that here we ignore abundancies of individual B-cell clonotypes and count each of them only once when summing over substitutions. This is reasonable as it removes substitution frequnecy biases coming from preferential expansion of B-cell clonotypes with certain hypermutations.

```{r echo = FALSE, include=FALSE, cache=FALSE}
library(seqLogo)
library(dplyr)
library(ggplot2)
library(reshape2)
library(stringr)
library(tidyr)
library(scales)
library(RColorBrewer)
library(gplots)
```

```{r, fig.height=10, echo = FALSE}
load('shm.Rda')
shm$region <- factor(shm$region, c("FR1","CDR1","FR2","CDR2","FR3","CDR3"))
shm$clonotypes <- 1
shm$name <- paste(shm$proj, shm$sample, shm$cells, sep='_')

contig.melted <- shm %>% dplyr::select(proj, sample, cells, name, contignt) %>%
  unique() %>%
  mutate(nt = str_split(contignt, '')) %>%
  unnest(nt = nt)

contig.total <- contig.melted %>% dplyr::group_by(proj, sample, cells, name) %>%
  dplyr::summarise(total.nt = n())

contig.freq <- contig.melted %>% dplyr::group_by(proj, sample, cells, nt, name) %>%
  dplyr::summarise(count = n()) %>%
  merge(contig.total) %>%
  mutate(nt.freq = count/total.nt)

total.mutations <- shm %>% dplyr::group_by(proj, sample, cells, name) %>%
  dplyr::summarise(total.mutations = n())

f <- shm %>% dplyr::group_by(proj, sample, cells, name, from, to) %>%
  dplyr::summarise(total.clonotypes = sum(clonotypes)) %>%
  merge(dplyr::select(contig.freq, from=nt, proj, sample, cells, nt.freq, name)) %>%
  merge(total.mutations) %>%
  mutate(weight = total.clonotypes/(total.mutations*nt.freq))
  
f1 <- f %>% dplyr::group_by(proj, sample, cells, name) %>%
  dplyr::summarise(weight.total = sum(weight))

f <- f %>% merge(f1) %>%
  mutate(freq = weight / weight.total) %>%
  dplyr::select(proj, sample, cells, from, to, freq, name)

st <- read.table('freq_steele.txt', header=T) %>%
  mutate(proj = 'steele', sample = 'steele', cells='N', name = 'steele',
         freq = freq/100)

f <- rbind(f, st)

f$from <- factor(f$from, levels = rev(c("A", "T", "C", "G")))
f$to <- factor(f$to, levels = c("A", "T", "C", "G"))

ggplot(f, aes(x = to, y = from)) +
  geom_tile(aes(fill = freq)) +
  geom_text(aes(label = percent(freq)), cex=3) +
  facet_wrap(~name, ncol=3) +
  scale_fill_gradient2("Frequency", low = "#F4F27B", mid = "#67CB87", high = "#2F6B89", midpoint = 0.1, na.value = "white")
```

We compute Jensen-Shannon divergences (``JSD``), a metric that is commonly used to compare frequency distributions. The smaller the divergence, the closer are substitution frequency distributions. Of note:

* Old donors appear to be outliers at this plot, but we cannot rule out batch effect in the absence of controls
* Samples for ``flu`` are highly correlated. Unfortunately this also includes control. All theese samples come from the same donor.

```{r, fig.width=6.5, fig.height=6, echo = FALSE}
jsd <- function(v1, v2){
  m <- 0.5 * (v1 + v1)
  0.5 * (sum(v1 * log2(v1 / m)) + sum(v2 * log2(v2 / m)))
}

f <- f[ order(f$from, f$to), ]

jsd.df <- data.frame()

for (i in unique(f$name)){
  x <- filter(f, name == i)$freq
  
  .jsd.df <- f %>% dplyr::group_by(name) %>%
  dplyr::summarise(jsd = jsd(x, freq))
  
  .jsd.df$x <- i

  jsd.df <- rbind(jsd.df, .jsd.df)
}

jsd.df$jsd2 = round(100 * jsd.df$jsd, 1)

ggplot(jsd.df %>% filter(!(x %in% c("raji_raji_P", "steele")) & !(name %in% c("raji_raji_P", "steele"))), 
       aes(x, name)) +
  geom_tile(aes(fill = jsd * 100)) +
  geom_text(aes(label = jsd2), cex=3) +
  xlab("") + ylab("") +
  scale_fill_gradientn("JSD, x100", colors=colorRampPalette(brewer.pal(11, 'Spectral'))(32), limits=c(1e-2, 2.7)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text.y = element_text(angle = 45, hjust = 1),
        legend.position="bottom")
```

Compare substitution frequency distributions of our samples with ``steele`` reference and ``raji`` cell line. Note that ``raji`` is an extreme outlier, this is quite obvious from the substitution frequency matrices given above. The data from ``steele`` is far more similar to our results, but still more than $2$ times farther in terms of ``JSD`` distance from each sample than the sample if from its most distant counterpart in our vaccinated donor set.

```{r, echo = FALSE}
ggplot(jsd.df %>% filter((x %in% c("raji_raji_P", "steele"))), 
       aes(x, name)) +
  geom_tile(aes(fill = jsd * 100)) +
  geom_text(aes(label = jsd2), cex=3) +
  scale_x_discrete("", expand = c(0,0)) + ylab("") +
  scale_fill_gradientn("JSD, x100", colors=colorRampPalette(brewer.pal(11, 'Spectral'))(32), limits=c(1e-2, 17.2))
```

Once more ``raji`` and ``steele`` substitution frequencies side-by-side. The ``C>>G`` rule does not hold for ``raji`` sample.

```{r, echo = FALSE}
ggplot(f %>% filter(name %in% c("steele", "raji_raji_P")), aes(x = to, y = from)) +
  geom_tile(aes(fill = freq)) +
  geom_text(aes(label = percent(freq)), cex=3) +
  facet_wrap(~name, ncol=3) +
  scale_fill_gradient2("Frequency", low = "#F4F27B", mid = "#67CB87", high = "#2F6B89", midpoint = 0.1, na.value = "white")
```

## Clustering samples based on mutation profile

```{r, fig.height=5, echo = FALSE, cache=FALSE}
f.mat <- f %>% dplyr::mutate(type = paste(from, to, sep = ">")) %>%
  dplyr::select(name, type, freq) %>%
  dcast(name ~ type)

rownames(f.mat) <- f.mat$name
f.mat$name <- NULL

heatmap.2(as.matrix(f.mat),
          hclustfun = function(x) hclust(x, method = "ward.D2"),
          trace = "none",
          col=colorRampPalette(brewer.pal(9, 'YlGn'))(32))
```

## Distribution of substitutions by position in template

```{r, fig.height=5, echo = FALSE}
shm$type <- with(shm, paste(from, to, sep = ">")) 

ggplot(shm %>% filter(!is.na(region)), aes(x=pos/nchar(contignt), color = type)) +
  geom_density() + 
  xlab("Relative position in the template") + ylab("Probability density") +
  facet_grid(from~to)
```

## Mining for ADAR and AID signatures

Let us first define a set of four variables corresponding to ADAR/AID signatures:

The fraction of mutations originating from a given base type $i$ is $f_{i\cdot} = \sum_{j\in A,T,G,C} f_{ij}$

- AID prevalence $AID_{p} = f_{C\cdot} + f_{G\cdot}$
- AID strand bias $AID_{s} = f_{G\cdot} / \left(f_{C\cdot} + f_{G\cdot}\right)$
- ADAR prevalence $ADAR_{p} = f_{A\cdot} + f_{T\cdot}$
- ADAR strand bias $ADAR_{s} = f_{A\cdot} / \left(f_{A\cdot} + f_{T\cdot}\right)$

The plot below shows the aforementioned values for each sample. Reference values for ``raji`` and ``steele`` are shown in ``red`` and ``blue`` respectively.

```{r, echo = FALSE, cache=FALSE}
f2 <- f %>% dplyr::group_by(proj, sample, cells, name) %>%
  dplyr::summarise(AIDp = sum(freq[which(from %in% c("C", "G"))]),
                   AIDs = sum(freq[which(from == "G")]) / AIDp,
                   ADARp = sum(freq[which(from %in% c("A", "T"))]),
                   ADARs = sum(freq[which(from == "A")]) / ADARp) %>%
  melt()

f2$metric = with(f2, ifelse(variable %in% c("AIDp", "ADARp"), "prevalence", "strand bias"))
f2$enzyme = with(f2, ifelse(variable %in% c("AIDp", "AIDs"), "AID", "ADAR"))

ggplot() +
  geom_point(data = subset(f2, !(name %in% c("steele", "raji_raji_P"))), aes(x=name, y = value), stat = "identity") +
  geom_hline(data = subset(f2, name == "raji_raji_P"), aes(yintercept = value), color = "red", linetype = "dashed") +
  geom_hline(data = subset(f2, name == "steele"), aes(yintercept = value), color = "blue", linetype = "dashed") +
  facet_grid(metric~enzyme, scales="free") +
  ylab("") + xlab("") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r, fig.width=4, echo = FALSE}
# f2 <- f %>% dplyr::group_by(proj, sample, cells, name) %>%
#   dplyr::summarise(freq = sum(freq)) %>%
#   dcast(proj + sample + cells +name ~ from, value.var='freq')
# 
# f2$AID_p = f2$A/(f2$A+f2$"T")
# f2$G.GC.ratio = f2$G/(f2$G+f2$C)
# f2 <- f2 %>%
#   dplyr::select(name, A.AT.ratio, G.GC.ratio) %>%
#   melt(id.vars=c('name'))
# 
# ggplot(f2, aes(variable, name)) +
#   geom_tile(aes(fill = value)) +
#   geom_text(aes(label = round(value, 3)), cex=3) +
#   scale_fill_gradient2(low = "#2F6B89", mid = "#67CB87", high = "#F4F27B", midpoint = 0.6, 
#                        na.value = "white")
```

```{r, fig.height=2, fig.width=4, echo = FALSE}
# shm$type <- paste(shm$from, shm$to, sep='>')
# 
# for (i in unique(shm$type)){
#   print(i)
#   set <- filter(shm, type == i, nchar(context) == 7, str_detect(context, 'N') == F) %>%
#     separate(context, c('n.1', 'n.2', 'n.3', 'n.4', 'n.5', 'n.6', 'n.7'), c(1,2,3,4,5,6)) %>%
#     dplyr::select(starts_with('n.'))
#   
#   prob <- data.frame()
#   
#   for (j in c('A','C','G','T')){
#     .prob <- apply(set, 2, function(x) length(x[x==j])/length(x))
#     prob <- rbind(prob, .prob)
#   }
#   rownames(prob) <- c('A','C','G','T')
#   colnames(prob) <- paste0('p', 1:7)
#   seqLogo(makePWM(prob), ic.scale=F,  xfontsize=10, yfontsize=10)
# }
```

```{r, fig.height=2, fig.width=4, echo = FALSE}
# shm$type <- paste(shm$from, shm$to, sep='>')
# 
# for (i in unique(shm$from)){
#   print(i)
#   set <- filter(shm, from == i, nchar(context) == 7, str_detect(context, 'N') == F) %>%
#     separate(context, c('n.1', 'n.2', 'n.3', 'n.4', 'n.5', 'n.6', 'n.7'), c(1,2,3,4,5,6)) %>%
#     dplyr::select(starts_with('n.'))
#   
#   set$n.4 = i
#   
#   prob <- data.frame()
#   
#   for (j in c('A','C','G','T')){
#     .prob <- apply(set, 2, function(x) length(x[x==j])/length(x))
#     prob <- rbind(prob, .prob)
#   }
#   rownames(prob) <- c('A','C','G','T')
#   colnames(prob) <- paste0('p', 1:7)
#   seqLogo(makePWM(prob), ic.scale=F,  xfontsize=10, yfontsize=10)
# }
```
